---
title: "SR2 Chapter 3 Hard"
author: "Brian Callander"
date: "2020-04-05"
tags: statistical rethinking, solutions, grid approximation, posterior predictive check, posterior predictive distribution, map, binomial, hpdi
tldr: "Here's my solution to the hard exercises in chapter 3 of McElreath's Statistical Rethinking, 2nd edition."
always_allow_html: yes
output: 
  md_document:
    variant: markdown
    preserve_yaml: yes
---

Here's my solutions to the hard exercises in chapter 3 of McElreath's Statistical Rethinking, 2nd edition. 

<!--more-->

<div> 
  $\DeclareMathOperator{\dbinomial}{Binomial}
   \DeclareMathOperator{\dbernoulli}{Bernoulli}
   \DeclareMathOperator{\dpoisson}{Poisson}
   \DeclareMathOperator{\dnormal}{Normal}
   \DeclareMathOperator{\dt}{t}
   \DeclareMathOperator{\dcauchy}{Cauchy}
   \DeclareMathOperator{\dexponential}{Exp}
   \DeclareMathOperator{\duniform}{Uniform}
   \DeclareMathOperator{\dgamma}{Gamma}
   \DeclareMathOperator{\dinvpamma}{Invpamma}
   \DeclareMathOperator{\invlogit}{InvLogit}
   \DeclareMathOperator{\logit}{Logit}
   \DeclareMathOperator{\ddirichlet}{Dirichlet}
   \DeclareMathOperator{\dbeta}{Beta}$
</div>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  error = TRUE,
  # cache = TRUE,
  dev = "svglite"
)

library(tidyverse)
library(scales)
library(kableExtra)

library(rethinking)

theme_set(theme_minimal())

set.seed(60326)
```

Let's first put the data into a tibble for easier manipulation later.

```{r data}
data(homeworkch3)

df <- tibble(birth1 = birth1, birth2 = birth2) %>% 
  mutate(birth = row_number())
```

```{r data_table, echo=F}
df %>% 
  head() %>% 
  kable(caption = "The first few rows of the data.") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"))
```

## 3H1

Let's check we have the correct total cound and the correct number of boys.

```{r}
h1_counts <- df %>% 
  gather(order, gender, -birth) %>% 
  summarise(boys = sum(gender), births = n())
```

Now we can grid approximate the posterior as before.

```{r h1_posterior}
granularity <- 1000

h1_grid <- tibble(p = seq(0, 1, length.out = granularity)) %>% 
  mutate(prior = 1)

h1_posterior <- h1_grid %>% 
  mutate(
    likelihood = dbinom(h1_counts$boys, h1_counts$births, p),
    posterior = prior * likelihood,
    posterior = posterior / sum(posterior)
  )
```

The maximum a posteriori (MAP) value is the value of p that maximises the posterior.

```{r h1}
h1_map <- h1_posterior %>% 
  slice(which.max(posterior)) %>% 
  pull(p)

h1_map
```


```{r h1_posterior_plot, echo=F, fig.cap="Solution 3H1: posterior probability of giving birth to a boy."}
h1_posterior %>% 
  ggplot() +
  aes(p, posterior) +
  geom_area(fill = 'skyblue', colour = 'white') +
  geom_vline(xintercept = h1_map, linetype = 'dashed', colour = 'chocolate', size = 1) +
  labs(
    x = 'Probability of a boy',
    y = 'Probability density',
    title = '3H1: Posterior probability of giving birth to a boy',
    subtitle = str_glue("The dashed line indicates the MAP = {signif(h1_map, 3)}")
  )

```

## 3H2

We draw samples with weight equalt to the posterior. We then apply the `HPDI` function to these samples, each time with a different width.

```{r h2}
h2_samples <- h1_posterior %>% 
  sample_n(10000, replace = TRUE, weight = posterior) %>% 
  pull(p)

h2_hpdi <- h2_samples %>% 
  crossing(prob = c(0.5, 0.89, 0.97)) %>% 
  group_by(prob) %>% 
  group_map(HPDI) 

h2_hpdi
```

## 3H3

The posterior predictive samples are possible observations according to our posterior.

```{r h3_posterior_predictive}
h3_posterior_predictive <- rbinom(10000, 200, h2_samples)
```


```{r h3_plot, fig.cap="Solution 3H3: the posterior predictive distribution for 200 births", echo = FALSE}
h3_posterior_predictive %>% 
  tibble(posterior_predictive = .) %>% 
  ggplot() +
  aes(posterior_predictive) +
  geom_histogram() +
  geom_vline(xintercept = h1_counts$boys, linetype = 'dashed', colour = 'chocolate', size = 1) +
  scale_y_continuous(labels = comma) +
  labs(
    title = '3H3: the posterior predictive distribution for 200 births',
    subtitle = "The dashed line indicates the observed number of boys in 200 births",
    x = 'Number of boys in 200 births',
    y = 'Proportion'
  )
```

The number of observed births is very close to the MAP of the posterior predictive distribution, suggesting we have a decent fit.

## 3H4

Our data are from birth pairs and so far we didn't make any distinction between the first and second births. To test this assumption, we can perform a posterior predictive check as in 3H3, but this time for first births.

```{r h4_posterior_predictive}
h4_posterior_predictive <- rbinom(10000, 100, h2_samples)
```


```{r h4_posterior_predictive_plot, fig.cap="Solution 3H4: the posterior predictive distribution for 100 births", echo = FALSE}
h4_posterior_predictive %>% 
  tibble(posterior_predictive = .) %>% 
  ggplot() +
  aes(posterior_predictive) +
  geom_histogram() +
  geom_vline(xintercept = sum(df$birth1), linetype = 'dashed', colour = 'chocolate', size = 1) +
  scale_y_continuous(labels = comma) +
  labs(
    title = '3H4: the posterior predictive distribution for 100 births',
    subtitle = "The dashed line indicates the observed number of boys in 100 first-born births",
    x = 'Number of boys in 100 births',
    y = 'Proportion'
  )
```

The fit doesn't look quite as good for first births as it did for all births together. It also doesn't look bad since there is still a fair bit of probability mass around the observed number of first birth boys.

## 3H5

As the final posterior predictive check, let's check the number of boys born after a girl.

```{r h5}
h5_counts <- df %>% 
  filter(birth1 == 0) %>% 
  summarise(boys = sum(birth2), births = n())

h5_posterior_predictive <- rbinom(10000, h5_counts$births, h2_samples)
```


```{r h5_posterior_predictive, fig.cap="Solution 3H5: the posterior predictive distribution for 100 births", echo = FALSE}
h5_posterior_predictive %>% 
  tibble(posterior_predictive = .) %>% 
  ggplot() +
  aes(posterior_predictive) +
  geom_histogram() +
  geom_vline(xintercept = h5_counts$boys, linetype = 'dashed', colour = 'chocolate', size = 1) +
  scale_y_continuous(labels = comma) +
  labs(
    title = str_glue('3H5: the posterior predictive distribution for {h5_counts$births} births'),
    subtitle = str_glue("The dashed line indicates the observed number of boys following the {h5_counts$births} first-born girls"),
    x = str_glue('Number of boys in {h5_counts$births} births'),
    y = 'Proportion'
  )
```

The fit here looks bad, since the observed number of boys is higher than the bulk of the model's expectations.
