---
title: "SR2 Chapter 2 Hard"
author: "Brian Callander"
date: "2020-03-01"
tags: statistical rethinking, solutions, conditional probability, counting, bayes rule, pandas
tldr: "Here's my solution to the hard exercises in chapter 2 of McElreath's Statistical Rethinking, 1st edition."
always_allow_html: yes
output: 
  md_document:
    variant: markdown
    preserve_yaml: yes
---

Here's my solution to the hard exercises in chapter 2 of McElreath's Statistical Rethinking, 1st edition. When writing this up, I came across a [very relevant article](https://www.theguardian.com/world/2020/feb/28/red-pandas-are-actually-two-separate-species-study-finds). We'll solve these problems in two ways: using the counting method and using Bayes rule.

<!--more-->

<div> 
  $\DeclareMathOperator{\dbinomial}{Binomial}
   \DeclareMathOperator{\dbernoulli}{Bernoulli}
   \DeclareMathOperator{\dpoisson}{Poisson}
   \DeclareMathOperator{\dnormal}{Normal}
   \DeclareMathOperator{\dt}{t}
   \DeclareMathOperator{\dcauchy}{Cauchy}
   \DeclareMathOperator{\dexponential}{Exp}
   \DeclareMathOperator{\duniform}{Uniform}
   \DeclareMathOperator{\dgamma}{Gamma}
   \DeclareMathOperator{\dinvpamma}{Invpamma}
   \DeclareMathOperator{\invlogit}{InvLogit}
   \DeclareMathOperator{\logit}{Logit}
   \DeclareMathOperator{\ddirichlet}{Dirichlet}
   \DeclareMathOperator{\dbeta}{Beta}$
</div>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  error = TRUE,
  # cache = TRUE,
  dev = "svglite"
)

library(tidyverse)
library(scales)
library(kableExtra)

library(rethinking)

theme_set(theme_minimal())

set.seed(97367)

```

## Counting method

Let's generate a dataset with all the features necessary to solve all the questions: twins at first birth, twins at second birth, and testing positive for species A.

```{r}
N <- 100000

dfa <- tibble(
    species = 'A',
    t1 = rbinom(N, 1, 0.1),
    t2 = rbinom(N, 1, 0.1),
    pa = rbinom(N, 1, 0.8)
  )

dfb <- tibble(
    species = 'B',
    t1 = rbinom(N, 1, 0.2),
    t2 = rbinom(N, 1, 0.2),
    pa = rbinom(N, 1, 1 - 0.65)
  )

df <- dfa %>% bind_rows(dfb)
```

All of the problems can now be solved by simply filtering out any events not consisent with our observations, then summarising the remaining events.

```{r}
h1 <- df %>% 
  filter(t1 == 1) %>% 
  summarise(mean(t2 == 1)) %>% 
  pull()

h2 <- df %>% 
  filter(t1 == 1) %>% 
  summarise(mean(species == 'A')) %>% 
  pull()

h3 <- df %>% 
  filter(t1 == 1, t2 == 0) %>% 
  summarise(mean(species == 'A')) %>% 
  pull()

h4a <- df %>% 
  filter(pa == 1) %>% 
  summarise(mean(species == 'A')) %>% 
  pull()

h4b <- df %>% 
  filter(pa == 1, t1 == 1, t2 == 0) %>% 
  summarise(mean(species == 'A')) %>% 
  pull()
```

```{r test, ref.label='solutions', echo=F}
```

For H1 we expect the probability to be between 0.1 and 0.2, since those are the two possible birth rates. Also, since we observed a twin birth already, it makes sense that it is closer to 0.2 since species B is more likely to birth twins. In other words, in H2 we expect the species to be less likely to be species A. Birthing a singleton infant is fairly common, so we wouldn't expect this observation to change our inference very much in H3.

## Bayes rule

Let's also work out the solutions analytically using Bayes rule. Let's start with H2 since it's useful for calculating H1.

$$
\begin{align}
  \mathbb P(A \mid T_1)
  &=
  \frac{\mathbb P(T_1 \mid A) \mathbb P(A)}{\mathbb P(T_1)}
  \\
  &=
  \frac{\mathbb P(T_1 \mid A) \mathbb P(A)}{\mathbb P(T_1 \mid A) \mathbb P(A) + \mathbb P(T_1 \mid B) \mathbb P(B)}
  \\
  &=
  \frac{0.1 \cdot 0.5}{0.1 \cdot 0.5 + 0.2 \cdot 0.5}
  \\
  &=
  \frac{0.05}{0.05 + 0.1}
  \\
  &=
  \frac{1}{3}
\end{align}
$$

Now we can use our solution to H2 and plug it into the appropriate place in the formula for H1. Note that $\mathbb P(T_2 \mid A)$ is the same as $\mathbb P(T_1 \mid A)$ by the assumptions of the problem. Similarily, once we know the species, whether the first birth was twins is irrelevant to the probability of twins in the second birth, i.e. $\mathbb P(T_2 \mid T_1, A) = \mathbb P(T_2 \mid A)$.

$$
\begin{align}
  \mathbb P(T_2 \mid T_1)
  &=
  \mathbb P(T_2 \mid T_1, A) \mathbb P(A \mid T_1)
  +
  \mathbb P(T_2 \mid T_1, B) \mathbb P(B \mid T_1)
  \\
  &=
  \mathbb P(T_2 \mid A) \mathbb P(A \mid T_1)
  +
  \mathbb P(T_2 \mid B) \mathbb P(B \mid T_1)
  \\
  &=
  \frac{1}{10} \cdot \frac{1}{3} + \frac{2}{10} \cdot \frac{2}{3}
  \\
  &=
  \frac{5}{30}
  \\
  &=
  \frac{1}{6}
\end{align}
$$

For H3, let's use the notation $-T_i$ to mean singleton infants (i.e. not twins).


$$
\begin{align}
  \mathbb P(A \mid T_1, - T_2)
  &=
  \frac{\mathbb P(- T_2 \mid T_1, A) \mathbb P(A \mid T_1)}{\mathbb P(- T_2 \mid T_1)}
  \\
  &=
  \frac{\mathbb P(- T_2 \mid A) \mathbb P(A \mid T_1)}{\mathbb P(- T_2 \mid T_1)}
  \\
  &=
  \frac{(1 - 0.1) \cdot \frac{1}{3}}{1 - 0.15}
  \\
  &=\frac{0.3}{0.85}
  \\
  &=
  \frac{6}{17}
\end{align}
$$

This is about `r signif(6 / 17, 3)`.

Now for H4a.

$$
\begin{align}
  \mathbb P(A \mid P_A)
  &=
  \frac{\mathbb P(P_A \mid A) \mathbb P(A)}{\mathbb P(P_A)}
  \\
  &=
  \frac{\mathbb P(P_A \mid A) \mathbb P(A)}{\mathbb P(P_A \mid A) \mathbb P(A) + \mathbb P(P_A \mid B) \mathbb P(B)}
  \\
  &=
  \frac{0.8 \cdot 0.5 }{0.8 \cdot 0.5 + 0.35 \cdot 0.5}
  \\
  &=
  \frac{0.4 }{0.4 + 0.175}
  \\
  &=
  \frac{0.4 }{0.575}
\end{align}
$$

This is about `r signif(0.4 / 0.575, 3)`.

Finally H4b.

$$
\begin{align}
  \mathbb P(A \mid P_A, T_1, -T_2)
  &=
  \frac{\mathbb P(P_A \mid A, T_1, -T_2) \mathbb P(A \mid T_1, -T_2)}{\mathbb P(P_A \mid T_1, -T_2)}
  \\
  &=
  \frac{\mathbb P(P_A \mid A) \mathbb P(A \mid T_1, -T_2)}{\mathbb P(P_A \mid A) \mathbb P(A \mid T_1, -T_2) + \mathbb P(P_A \mid B) \mathbb P(B \mid T_1, -T_2)}
  \\
  &=
  \frac{\frac{4}{5} \cdot \frac{6}{17} }{\frac{4}{5}\cdot \frac{6}{17} + \frac{7}{20} \cdot \frac{11}{17}}
  \\
  &=
  \frac{\frac{24}{85} }{\frac{24}{85} + \frac{77}{340}}
  \\
  &=
  \frac{\frac{24}{85} }{\frac{92 + 77}{340}}
  \\
  &=
  \frac{24}{85} \cdot \frac{340}{169}
  \\
  &=
  \frac{92}{169} 
\end{align}
$$

This is about `r signif(92 / 169, 3)`.

```{r solutions, echo=FALSE, include=FALSE}
tibble(
    h1_counting = h1,
    h1_bayes = 1/6,
    h2_counting = h2,
    h2_bayes = 1 / 3,
    h3_counting = h3,
    h3_bayes = 6 / 17,
    h4a_counting = h4a,
    h4a_bayes = 0.4 / 0.575,
    h4b_counting = h4b,
    h4b_bayes = 92 / 169,
  ) %>% 
  gather(exercise, solution) %>% 
  mutate(
    method = exercise %>% str_extract("([^_]+)$"),
    exercise = exercise %>% str_replace("_.*", "")
  ) %>% 
  spread(method, solution) %>% 
  kable(caption = 'Solutions') %>% 
  kable_styling(bootstrap_options = c('hover', 'striped', 'responsive'))
```
