---
title: "SR2 Chapter 3 Medium"
author: "Brian Callander"
date: "2020-04-05"
tags: statistical rethinking, solutions, grid approximation, posterior probability, posterior predictive probability, hpdi, binomial
tldr: "Here's my solution to the medium exercises in chapter 3 of McElreath's Statistical Rethinking, 2nd edition."
always_allow_html: yes
output: 
  md_document:
    variant: markdown
    preserve_yaml: yes
---

Here's my solution to the medium exercises in chapter 3 of McElreath's Statistical Rethinking, 2nd edition. 

<!--more-->

<div> 
  $\DeclareMathOperator{\dbinomial}{Binomial}
   \DeclareMathOperator{\dbernoulli}{Bernoulli}
   \DeclareMathOperator{\dpoisson}{Poisson}
   \DeclareMathOperator{\dnormal}{Normal}
   \DeclareMathOperator{\dt}{t}
   \DeclareMathOperator{\dcauchy}{Cauchy}
   \DeclareMathOperator{\dexponential}{Exp}
   \DeclareMathOperator{\duniform}{Uniform}
   \DeclareMathOperator{\dgamma}{Gamma}
   \DeclareMathOperator{\dinvpamma}{Invpamma}
   \DeclareMathOperator{\invlogit}{InvLogit}
   \DeclareMathOperator{\logit}{Logit}
   \DeclareMathOperator{\ddirichlet}{Dirichlet}
   \DeclareMathOperator{\dbeta}{Beta}$
</div>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  error = TRUE,
  # cache = TRUE,
  dev = "svglite"
)

library(tidyverse)
library(scales)
library(kableExtra)

library(rethinking)

theme_set(theme_minimal())

set.seed(97367)
```

Assuming Earth has 70% water cover, and we observe water 8 times out of 15 globe tosses, let's calculate some posterior quantities with two choices of prior: uniform and step.

```{r}
p_true <- 0.7

W <- 8
N <- 15

granularity <- 1000 # points on the grid
```

## Uniform Prior

We calculate the grid approximation of the posterior as shown in the book.

```{r m1}
m1_grid <- tibble(p = seq(0, 1, length.out = granularity)) %>% 
  mutate(prior = 1)

m1_posterior <- m1_grid %>% 
  mutate(
    likelihood = dbinom(W, N, p),
    posterior = prior * likelihood
  )
```

```{r m1_plot, echo=F, fig.cap="Solution to exercise 3M1"}
m1_posterior %>% 
  ggplot() +
  aes(p, posterior) +
  geom_area(fill = 'skyblue', colour = 'white') +
  geom_vline(xintercept = p_true, colour = 'chocolate', linetype = 'dashed') +
  scale_x_continuous(labels = percent) +
  labs(
    x = "Water cover",
    y = "Probability",
    title = "3M1: Posterior probability of Earth's water cover",
    subtitle = paste(
      "Dashed line indicates true value",
      "Uniform prior",
      sep = "\n"
    )
  )
```

We can get draws from our posterior by sampling the water cover values many times with replacement, each value being drawn in proportion to the posterior probability. We can then just summarise these draws to get the desired interval.

```{r m2}
m2_samples <- m1_posterior %>% 
  sample_n(10000, replace = T, weight = posterior)

m2_hpdi <- HPDI(m2_samples$p, prob = 0.9)
m2_hpdi
```

The histogram looks as follows. This is much the same as the previous graph, but calculated from the samples.

```{r m2_plot, echo=F, fig.cap="Solution to exercise 3M2"}
m2_samples %>% 
  ggplot() +
  aes(p, fill = between(p, m2_hpdi[1], m2_hpdi[2])) + 
  geom_histogram() +
  geom_vline(xintercept = p_true, colour = 'black', linetype = 'dashed') +
  # geom_rect(aes(xmin = m2_hpdi[1], xmax = m2_hpdi[2], ymin = -20, ymax = 20), fill = "skyblue") +
  scale_x_continuous(labels = percent) +
  labs(
    x = "Water cover",
    y = "Number of draws",
    fill = "Within HPDI?",
    title = "3M2: Posterior draws",
    subtitle = paste(
      "Dashed line indicates true value",
      "Uniform prior",
      sep = "\n"
    )
  )
```

To get the posterior predictive sample, we take our posterior draws of $p$, then use them to draw a random number of observed water tosses out of 15. The fraction of posterior predictive samples with a given value is then the posterior predictive probability of that value.

```{r m3}
m3_prob <- m2_samples %>% 
  mutate(W = rbinom(n(), 15, p)) %>% 
  group_by(W) %>% 
  tally() %>% 
  mutate(probability = n / sum(n))
```


```{r m3_plot, echo=F, fig.cap="Solution to exercise 3M3"}
m3_prob %>% 
  ggplot() +
  aes(W, probability) +
  geom_col() +
  scale_x_continuous(breaks = seq(0, 16, 2)) +
  labs(
    x = "Number of tosses landing on water",
    y = "Posterior predictive probability",
    title = "3M3: Posterior predictive probability",
    subtitle = paste(
      "Uniform prior",
      sep = "\n"
    )
  )
```

We can also calculate the posterior predictive probabilities with a different number of tosses. Here with 9 tosses.

```{r m4}
m4_prob <- m2_samples %>% 
  mutate(W = rbinom(n(), 9, p)) %>% 
  group_by(W) %>% 
  tally() %>% 
  mutate(probability = n / sum(n))
```


```{r m4_plot, echo=F, fig.cap="Solution to exercise 3M4"}
m4_prob %>% 
  ggplot() +
  aes(W, probability) +
  geom_col() +
  scale_x_continuous(breaks = seq(0, 16, 1)) +
  labs(
    x = "Number of tosses landing on water",
    y = "Posterior predictive probability",
    title = "3M4: Posterior predictive probability",
    subtitle = paste(
      "Uniform prior",
      sep = "\n"
    )
  )
```

## Step Prior

Now we repeat the same steps but with the step prior instead of the uniform prior. We'll just repeat it without comment.

```{r m5_1}
m5_grid <- m1_grid %>% 
  mutate(prior = if_else(p < 0.5, 0, 1))

m5_posterior <- m5_grid %>% 
  mutate(
    likelihood = dbinom(W, N, p),
    posterior = prior * likelihood
  )
```

```{r m5_1_plot, echo=F, fig.cap="Solution to exercise 3M5 part 1"}
m5_posterior %>% 
  ggplot() +
  aes(p, posterior) +
  geom_area(fill = 'skyblue', colour = 'white') +
  geom_vline(xintercept = p_true, colour = 'chocolate', linetype = 'dashed') +
  scale_x_continuous(labels = percent) +
  labs(
    x = "Water cover",
    y = "Probability",
    title = "3M5 Part 1: Posterior probability of Earth's water cover",
    subtitle = paste(
      "Dashed line indicates true value",
      "Step prior",
      sep = "\n"
    )
  )
```

```{r m5_2}
m5_samples <- m5_posterior %>% 
  sample_n(10000, replace = T, weight = posterior)

m5_hpdi <- HPDI(m5_samples$p, prob = 0.9)
m5_hpdi
```

```{r m5_2_plot, echo=F, fig.cap="Solution to exercise 3M5 part 2"}
m5_samples %>% 
  ggplot() +
  aes(p, fill = between(p, m5_hpdi[1], m5_hpdi[2])) + 
  geom_histogram()  +
  geom_vline(xintercept = p_true, linetype = 'dashed') +
  scale_x_continuous(labels = percent) +
  labs(
    x = "Water cover",
    y = "Number of draws",
    fill = "Within HPDI?",
    title = "3M5 Part 2: Posterior draws",
    subtitle = paste(
      "Dashed line indicates true value",
      "Step prior",
      sep = "\n"
    )
  )
```

```{r m5_3}
m5_prob <- m5_samples %>% 
  mutate(W = rbinom(n(), 15, p)) %>% 
  group_by(W) %>% 
  tally() %>% 
  mutate(probability = n / sum(n))
```


```{r m5_3_plot, echo=F, fig.cap="Solution to exercise 3M5 part 3"}
m5_prob %>% 
  ggplot() +
  aes(W, probability) +
  geom_col() +
  scale_x_continuous(breaks = seq(0, 16, 1)) +
  labs(
    x = "Number of tosses landing on water",
    y = "Posterior predictive probability",
    title = "3M5 Part 3: Posterior predictive probability",
    subtitle = paste(
      "Step prior",
      sep = "\n"
    )
  )
```

```{r m5_4}
m5_prob <- m5_samples %>% 
  mutate(W = rbinom(n(), 9, p)) %>% 
  group_by(W) %>% 
  tally() %>% 
  mutate(probability = n / sum(n))
```


```{r m5_4_plot, echo=F, fig.cap="Solution to exercise 3M5 part 4"}
m5_prob %>% 
  ggplot() +
  aes(W, probability) +
  geom_col() +
  scale_x_continuous(breaks = seq(0, 16, 1)) +
  labs(
    x = "Number of tosses landing on water",
    y = "Posterior predictive probability",
    title = "3M5 Part 4: Posterior predictive probability",
    subtitle = paste(
      "Step prior",
      sep = "\n"
    )
  )
```

Let's compare the proportion of samples within 0.05 of the true value for each prior.

```{r prior_comparison}
p_close_uniform <- m2_samples %>% 
  group_by(close = p %>% between(p_true - 0.05, p_true + 0.05)) %>% 
  tally() %>% 
  mutate(probability = n / sum(n)) %>% 
  filter(close) %>% 
  pull(probability)

p_close_step <- m5_samples %>% 
  group_by(close = p %>% between(p_true - 0.05, p_true + 0.05)) %>% 
  tally() %>% 
  mutate(probability = n / sum(n)) %>% 
  filter(close) %>% 
  pull(probability)
```

The probability of being close to the true value under the uniform and step priors is `r p_close_uniform` and `r p_close_step`, respectively. The step prior thus has more mass around the true value.

## Exercise 3M6

Bayesian models are generative, meaning we can simulate new datasets according to our prior probabilities. We'll simulate 10 datasets for each value of N of interest. We simulate a dataset by randomly choosing a `p_true` from our uniform prior, then randomly choosing a `W` from the corresponding binomial distribution.

```{r m6_prior_predictive}
m6_prior_predictive <- crossing(
    N = 200 * (1:16), 
    iter = 1:10
  ) %>% 
  mutate(
    p_true = runif(n(), min=0, max=1), 
    W = rbinom(n(), N, p_true)
  )
```

For each of these simulated datasets, we grid approximate the posterior, take posterior samples, then calculate the HPDI.

```{r m6_hpdi}
m6_grid <- tibble(p = seq(0, 1, length.out = granularity)) %>% 
  mutate(prior = 1)

m6_posteriors <- m6_prior_predictive %>% 
  crossing(m6_grid) %>% 
  group_by(N, p_true, iter) %>% 
  mutate(
    likelihood = dbinom(W, N, p),
    posterior = prior * likelihood
  )

m6_samples <- m6_posteriors %>% 
  sample_n(1000, replace = TRUE, weight = posterior) 

m6_hpdi <- m6_samples %>% 
  summarise(lo = HPDI(p, 0.99)[1], hi = HPDI(p, 0.99)[2]) %>% 
  mutate(width = abs(hi - lo))
```

Now for each value of N, we check how many of the intervals have the desired width.

```{r m6_sample_size}
m6_n <- m6_hpdi %>% 
  group_by(N) %>% 
  summarise(fraction = mean(width < 0.05)) 
```


```{r m6_sample_size_plot, echo=F, fig.cap="Solution to exercise 3M6"}
m6_n %>% 
  ggplot() +
  aes(N, fraction) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = percent) +
  labs(
    x = "Sample size",
    y = "Fraction of datasets with sufficient precision",
    title = "3M6: Sample size estimation"
  )
```

Thus we expect a sample size around 2600-3000 to give us a sufficiently precise posterior estimation.
