---
title: "CIS Primer Question 2.3.1"
author: "Brian Callander"
date: "2019-01-26"
tags: CISP chapter 2, casual inference, dag, conditional independence
tldr: Here's my solution to question 2.3.1 from a Primer in Causal Inference in Statistics
always_allow_html: yes
output: 
  md_document:
    variant: markdown
    preserve_yaml: yes
---

Here are my solutions to Causal Inference in Statistics: a Primer (CISP), question 2.3.1.

<!--more-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  error = TRUE,
  # cache = TRUE,
  dev = "svglite"
)

library(tidyverse)
library(scales)
library(kableExtra)

theme_set(theme_bw())
 
```

We'll use the following dataset generated with the structure of figure 2.6 to check our answers. Note that the functions have been chosen so that all variables have approximately unit variance.

```{r}
set.seed(844909)

N <- 100000

df <- tibble(
  x = rnorm(N, 0, 1),
  r = x + rnorm(N, 0, 0.1),
  s = r + rnorm(N, 0, 0.1),
  v = rnorm(N, 0, 1),
  y = v + rnorm(N, 0, 0.1),
  u = v + rnorm(N, 0, 0.1),
  t = sqrt(0.5) * s + sqrt(0.5) * u + rnorm(N, 0, 0.1),
  p = t + rnorm(N, 0, 0.1)
)
```

## Part a

Conditional on the set $\{R, V\}$, the following pairs of variables are independent:

* $X, S$
* $X, T$
* $X, U$
* $X, Y$
* $S, U$
* $S, Y$
* $T, Y$
* $U, Y$

We can verify that this is the case for our simulated dataset by regressing the left of each listed pair on the right, including R and V as covariates. If the pair is independent, then the coefficient of the right variable should be around 0.

```{r part_a}
part_a <- list(
    xs = formula(x ~ 1 + r + v + s),
    xt = formula(x ~ 1 + r + v + t),
    xu = formula(x ~ 1 + r + v + u),
    xy = formula(x ~ 1 + r + v + y),
    su = formula(s ~ 1 + r + v + u),
    sy = formula(s ~ 1 + r + v + y),
    ty = formula(t ~ 1 + r + v + y),
    uy = formula(u ~ 1 + r + v + y)
  ) %>% 
  map(lm, df) %>% 
  map_dfr(broom::tidy, .id = 'model') %>% 
  filter(!(term %in% c('(Intercept)', 'r', 'v'))) %>%
  transmute(
    model,
    term,
    lower = estimate - 2 * std.error, 
    estimate,
    upper = estimate + 2 * std.error
  ) 
```

```{r part_a_plot, echo=FALSE, fig.cap='Variables that are independent given R and V'}
part_a %>% 
  arrange(desc(term), model) %>% 
  ggplot() +
  aes(x = model, ymin = lower, y = estimate, ymax = upper) +
  geom_hline(yintercept = 0, linetype = 'dashed', colour = 'chocolate') +
  geom_pointrange(size = 0.3) +
  labs(
    x = 'Variable pairs (A, B)',
    y = 'Coefficient estimate',
    title = 'Coefficient of B in the model lm(A ~ 1 + B + R + V)'
  ) +
  coord_flip()

```

## Part b

Given $R$, the following variables are independent:

* $X, S$
* $X, T$
* $X, U$
* $X, V$
* $X, Y$

Given $S$, the following variables are independent:

* $R, T$
* $R, U$
* $R, V$
* $R, Y$

The following variables are unconditionally independent:

* $S, U$
* $S, V$
* $S, Y$

Given $U$, the following variables are independent:

* $T, V$
* $T, Y$

Given $V$, the variables $U, Y$ are independent.

The above statements of independence can be verified in the same way as in part a.

```{r part_b}
part_b <- list(
    xs_given_r = formula(x ~ 1 + s + r),
    xt_given_r = formula(x ~ 1 + t + r),
    xu_given_r = formula(x ~ 1 + u + r),
    xv_given_r = formula(x ~ 1 + v + r),
    xy_given_r = formula(x ~ 1 + y + r),
    rt_given_s = formula(r ~ 1 + r + s),
    ru_given_s = formula(r ~ 1 + u + s),
    rv_given_s = formula(r ~ 1 + v + s),
    ry_given_s = formula(r ~ 1 + y + s),
    su = formula(s ~ 1 + u),
    sv = formula(s ~ 1 + v),
    sy = formula(s ~ 1 + y),
    tv_given_u = formula(t ~ 1 + v + u),
    ty_given_u = formula(t ~ 1 + v + u),
    uy_given_v = formula(u ~ 1 + y + v)
  ) %>% 
  map(lm, df) %>% 
  map_dfr(broom::tidy, .id = 'model') %>% 
  filter(term != '(Intercept)') %>% 
  transmute(
    model,
    term,
    lower = estimate - 2 * std.error, 
    estimate,
    upper = estimate + 2 * std.error
  ) 
```

```{r part_b_plot, echo=FALSE, fig.cap='...'}
part_b %>% 
  filter(upper < 0.5) %>% 
  ggplot() +
  aes(x = model, ymin = lower, y = estimate, ymax = upper, colour = term) +
  geom_hline(yintercept = 0, linetype = 'dashed') +
  geom_pointrange(size = 0.3) +
  labs(
    x = 'Conditional independence',
    y = 'Coefficient estimate',
    title = 'Conditional independence statements',
    colour = 'Term'
  ) +
  coord_flip() +
  # facet_grid(model ~ term) +
  NULL

```

## Part c

Given $R, P$, the following pairs are independent: 

* $X, S$
* $X, T$
* $X, U$
* $X, V$
* $X, Y$

```{r part_c}
part_c <- list(
    xs = formula(x ~ 1 + r + p + s),
    xt = formula(x ~ 1 + r + p + t),
    xu = formula(x ~ 1 + r + p + u),
    xv = formula(x ~ 1 + r + p + v),
    xy = formula(x ~ 1 + r + p + y)
  ) %>% 
  map(lm, df) %>% 
  map_dfr(broom::tidy, .id = 'model') %>% 
  filter(term != '(Intercept)', term != 'r', term != 'p') %>% 
  transmute(
    model,
    term,
    lower = estimate - 2 * std.error, 
    estimate,
    upper = estimate + 2 * std.error
  ) 
```

```{r part_c_plot, echo=FALSE, fig.cap='...'}
part_c %>% 
  filter(upper < 0.5) %>% 
  ggplot() +
  aes(x = model, ymin = lower, y = estimate, ymax = upper) +
  geom_hline(yintercept = 0, linetype = 'dashed', colour = 'chocolate') +
  geom_pointrange(size = 0.3) +
  labs(
    x = 'Pairs',
    y = 'Coefficient estimate',
    title = 'Conditional independence'
  ) +
  coord_flip() +
  # facet_grid(model ~ term) +
  NULL

```

## Part d

All statements from part b still hold. Moreover,

* given $R$, $(X, P)$ is independent;
* given $S$, $(R, P)$ is independent; and
* given $T$, $(S, P)$ is independent.

```{r part_d}
part_d <- list(
    xp_given_r = formula(x ~ 1 + p + r),
    rp_given_s = formula(r ~ 1 + p + s),
    sp_given_t = formula(s ~ 1 + p + t)
  ) %>% 
  map(lm, df) %>% 
  map_dfr(broom::tidy, .id = 'model') %>% 
  filter(term == 'p') %>% 
  transmute(
    model,
    term,
    lower = estimate - 2 * std.error, 
    estimate,
    upper = estimate + 2 * std.error
  ) 
```

```{r part_d_plot, echo=FALSE, fig.cap='...'}
part_d %>% 
  filter(upper < 0.5) %>% 
  ggplot() +
  aes(x = model, ymin = lower, y = estimate, ymax = upper) +
  geom_hline(yintercept = 0, linetype = 'dashed', colour = 'chocolate') +
  geom_pointrange(size = 0.3) +
  labs(
    x = 'Conditional independence',
    y = 'Coefficient estimate',
    title = 'Conditional independence statements',
    colour = 'Term'
  ) +
  coord_flip() +
  NULL

```

## Part e

The variables $X$ and $Y$ are independent given $Z$ if $Z$ is 

* $\emptyset$, 
* $\{ R \}$, 
* $\{ S \}$, 
* $\{ U \}$, 
* $\{ V \}$, 
* $\{ R, T \}$, 
* $\{ S, T \}$, 
* $\{ U, T \}$, 
* $\{ V, T \}$,

or any union of the above.

```{r part_e}
part_e <- list(
    empty = formula(y ~ 1 + x),
    r = formula(y ~ 1 + x + r),
    s = formula(y ~ 1 + x + s),
    u = formula(y ~ 1 + x + u),
    v = formula(y ~ 1 + x + v),
    rt = formula(y ~ 1 + x + r + t),
    st = formula(y ~ 1 + x + s + t),
    ut = formula(y ~ 1 + x + u + t),
    vt = formula(y ~ 1 + x + v + t)
  ) %>% 
  map(lm, df) %>% 
  map_dfr(broom::tidy, .id = 'model') %>% 
  filter(term == 'x') %>%
  transmute(
    model,
    term,
    lower = estimate - 2 * std.error, 
    estimate,
    upper = estimate + 2 * std.error
  ) 
```

```{r part_e_plot, echo=FALSE, fig.cap='Variables that make Y conditionally independent of X'}
part_e %>% 
  arrange(desc(term), model) %>% 
  ggplot() +
  aes(x = model, ymin = lower, y = estimate, ymax = upper) +
  geom_hline(yintercept = 0, linetype = 'dashed', colour = 'chocolate') +
  geom_pointrange(size = 0.3) +
  labs(
    x = 'Z variables',
    y = 'Estimate',
    title = 'Coefficient of X in the model lm(Y ~ 1 + X + Z)'
  )

```

## Part f

Conditioning on $T$ blocks all paths between $Y$ and $P$, so $P$ will have zero coefficient. On the other hand, this conditioning unblocks the path between $Y$ and $S$, so $S$ will have a non-zero coefficient. The conditioning on $S$ blocks the path from $X$ or $R$ to $Y$, so the coefficients of $X$ and $R$ should be zero.

```{r part_f}
part_f <- lm(y ~ 1 + x + r + s + t + p, df) %>% 
  broom::tidy() %>% 
  transmute(
    term,
    lower = estimate - 2 * std.error, 
    estimate, 
    upper = estimate + 2 * std.error
  ) 

```

```{r part_f_plot, echo=FALSE, fig.cap="Coefficients in the model Y ~ X + R + S + T + P"}
part_f %>% 
  ggplot() +
  aes(x = term, ymin = lower, y = estimate, ymax = upper) +
  geom_hline(yintercept = 0, linetype = 'dashed', colour = 'chocolate') +
  geom_pointrange(size = 0.3) +
  labs(
    x = 'Variable',
    y = 'Coefficient',
    title = 'Coefficients in the model lm(Y ~ 1 + X + R + S + T + P)'
  ) +
  coord_flip()

```

