---
title: "CIS Primer Question 3.4.1"
author: "Brian Callander"
date: "2019-02-15"
tags: CISP chapter 3, solutions, front door criteria, front door adjustment
tldr: Here are my solutions to question 3.4.1 of Causal Inference in Statistics a Primer (CISP).
always_allow_html: yes
output: 
  md_document:
    variant: markdown
    preserve_yaml: yes
---

Here are my solutions to question 3.4.1 of Causal Inference in Statistics: a Primer (CISP). 
$\DeclareMathOperator{\do}{do}$

<!--more-->



```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  error = TRUE,
  # cache = TRUE,
  dev = "svglite"
)

library(tidyverse)
library(scales)
library(kableExtra)
library(colorspace)
library(broom)

theme_set(theme_bw())
 
```

```{r functions, include=FALSE}
inv_logit <- function(x) 1 / (1 + exp(-x))
```

If we can only measure one additional variable to estimate the causal effect of $X$ on $Y$ in figure 3.8, then we should measure $W$. From [question 3.3.1](question_3_3_1.html) we see that no single variable satisfies the backdoor criteria. Moreover, visual inspection of the graph verifies that $W$ satisfies the frontdoor criteria:

1. it intercepts all (the only) directed paths from $X$ to $Y$;
2. there is no unblocked path from $X$ to $W$; and
3. all backdoor paths from $W$ to $Y$ are blocked by $X$.

To illustrate this, lets simulate the causal effect in 3 separate ways:

1. by intervention,
2. via the backdoor, and
3. via the frontdoor.

Here are the data. Note that we have created functions for $W$ and $Y$ for use later.

```{r data}
N <- 100000

W <- function(x) {
  N <- length(x)
  rbinom(N, 1, inv_logit(-x))
}

Y <- function(d, w, z) {
  N <- length(d)
  rbinom(N, 1, inv_logit(-d - w + 3*z))
}

df <- tibble(id = 1:N) %>% 
  mutate(
    b = rnorm(N, 0, 1),
    a = b + rnorm(N, 0, 0.1),
    c = rnorm(N, 0, 1),
    d = rbinom(N, 1, inv_logit(-1 + c)),
    z = rbinom(N, 1, inv_logit(-2 + 2*b + c)),
    x = rbinom(N, 1, inv_logit(a + z)),
    w = W(x),
    y = Y(d, w, z)
  )
```

```{r data_table, echo=FALSE}
df %>% 
  head(5) %>% 
  kable(caption = 'Simulated data for figure 3.8') %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"))
```

## Intervention

In order to simulate an intervention, we assign values to $X$ randomly, then assign new values for all its descendents. After intervention, the causal effect of $X$ on $Y$ is simply $\mathbb P(Y \mid X)$.

```{r intervention}
intervention <- df %>% 
  # intervene on x
  mutate(
    x = rbinom(n(), 1, 0.5),
    w = W(x),
    y = Y(d, w, z)
  ) %>% 
  # model P(y | do(x))
  glm(
    formula = y ~ x, 
    family = binomial(), 
    data = .
  ) %>% 
  # predict
  augment(
    newdata = tibble(x = 0:1), 
    type.predict = 'response'
  ) 

```

```{r intervention_table, echo=FALSE}
intervention %>% 
  kable(caption = 'P(Y | do(X))') %>% 
  kable_styling(bootstrap_options = c("hover", "responsive"))
```

We can compare this causal effect to the simple statistical effect to see the difference.

```{r noncausal}
noncausal <- df %>% 
  # model P(y | x)
  glm(
    formula = y ~ x, 
    family = binomial(), 
    data = .
  ) %>% 
  # predict
  augment(
    newdata = tibble(x = 0:1), 
    type.predict = 'response'
  ) 

```

```{r noncausal_table, echo=FALSE}
noncausal %>% 
  kable(caption = 'P(Y | X) ≠ P(Y | do(X))') %>% 
  kable_styling(bootstrap_options = c("hover", "responsive"))
```

## Backdoor

Since $\{X, Z\}$ satisfies the backdoor criteria, we can use it to apply the backdoor adjustment. First we'll need $\mathbb P(D, Z)$.

```{r p_d_z}
# P(d, z)
p_d_z <- df %>% 
  group_by(d, z) %>% 
  count() %>% 
  ungroup() %>%  
  mutate(p_d_z = n / sum(n)) 
```

Now we model $\mathbb P(Y \mid X, D, Z)$, multiply it by $\mathbb P(D, Z)$, then take the sum for each value of $X$.

```{r backdoor}  
backdoor <- formula(y ~ 1 + x + z + d) %>% 
  # model P(y | x, d, z)
  glm(
    family = binomial(),
    data = df
  ) %>%  
  # predict
  augment(
    type.predict = 'response',
    newdata = 
      crossing(
        d = c(0, 1),
        x = c(0, 1),
        z = c(0, 1)
      )
  ) %>% 
  # get P(d, z)
  mutate(p_y_given_d_x_z = .fitted) %>% 
  inner_join(p_d_z, by = c('d', 'z')) %>% 
  # backdoor adjustment over d, z
  group_by(x) %>% 
  summarise(p_y_given_do_x = sum(p_y_given_d_x_z * p_d_z))

```

```{r backdoor_table, echo=FALSE}
backdoor %>% 
  kable(caption = 'Backdoor estimates for P(Y | do(X))') %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"))
```

Note that the backdoor adjusted estimates are similar to the estimates from intervention.

## Frontdoor

To apply the frontdoor adjustment with $W$, we'll need $\mathbb P(W \mid X)$, $\mathbb P(X^\prime)$, and $\mathbb P(Y \mid X, W)$.

```{r probs}
p_w_given_x <- df %>% 
  group_by(x, w) %>% 
  count() %>% 
  group_by(x) %>% 
  mutate(p_w_given_x = n / sum(n)) %>% 
  ungroup()

p_xprime <- df %>% 
  group_by(xprime = x) %>% 
  count() %>% 
  ungroup() %>% 
  mutate(p_xprime = n / sum(n))

p_y_given_xprime_w <- formula(y ~ 1 + x + w) %>% 
  glm(
    family = binomial(),
    data = df
  ) %>% 
  augment(
    newdata = crossing(x = 0:1, w = 0:1),
    type.predict = 'response'
  ) %>% 
  transmute(
    xprime = x,
    w,
    p_y_given_xprime_w = .fitted
  )
```

Now we apply the frontdoor adjustment:

$$
\mathbb P (Y \mid \do(X))
=
\sum_{x^\prime, w}
\mathbb P(x^\prime)
\cdot
\mathbb P(w \mid x)
\cdot
\mathbb P (y \mid x^\prime, w)
.
$$

```{r frontdoor}
frontdoor <- p_w_given_x %>% 
  inner_join(p_y_given_xprime_w, by = 'w') %>% 
  inner_join(p_xprime, by = 'xprime') %>% 
  group_by(x) %>%
  summarise(sum(p_w_given_x * p_y_given_xprime_w * p_xprime))

```


```{r frontdoor_table, echo=FALSE}
frontdoor %>% 
  kable(caption = 'Frontdoor estimates of P(Y | do(X))') %>% 
  kable_styling(bootstrap_options = c("hover", "responsive"))
```

Our frontdoor estimates  of $\mathbb P(Y \mid \do(X))$ are very similar to the intervention and backdoor estimates.
