---
title: "CIS Primer Question 3.2.1"
author: "Brian Callander"
date: "2019-02-10"
tags: CISP chapter 3, solutions, ace, risk difference, simpsons reversal
tldr: Here are my solutions to question 3.2.1 of Causal Inference in Statistics a Primer (CISP).
always_allow_html: yes
output: 
  md_document:
    variant: markdown
    preserve_yaml: yes
---

Here are my solutions to question 3.2.1 of Causal Inference in Statistics: a Primer (CISP). 

<!--more-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  error = TRUE,
  # cache = TRUE,
  dev = "svglite"
)

library(tidyverse)
library(scales)
library(kableExtra)
library(colorspace)

theme_set(theme_bw())
 
```

Here are the parameters we'll use. Note that they are taken from the Simpson's revesal example of [question 1.5.2](question_1_5_2.html).

```{r params}
r	<- 0.28   # fraction with syndrome

q0 <- 0.07  # P(X = 1 | Z = 0)
q1 <- 0.85  # P(X = 1 | Z = 1)
             
p00 <- 0.84 # P(Y = 1 | X = 0, Z = 0)
p10 <- 0.88 # P(Y = 1 | X = 1, Z = 0)
p01 <- 0.53 # P(Y = 1 | X = 0, Z = 1)
p11 <- 0.58 # P(Y = 1 | X = 1, Z = 1)
```

## Part a

We can simulate the intervention by generating values for $X$ independently of $Z$.

```{r data_a}
N <- 10000  # number of individuals

set.seed(53201)

part_a <- tibble(z = rbinom(N, 1, r)) %>% 
  mutate(
    x = rbinom(n(), 1, 0.5), # no Z-dependence
    p_y_given_x_z = case_when(
      x == 0 & z == 0 ~ p00,
      x == 0 & z == 1 ~ p01,
      x == 1 & z == 0 ~ p10,
      x == 1 & z == 1 ~ p11
    ),
    y = rbinom(n(), 1, p_y_given_x_z)
  ) %>% 
  group_by(x, y) %>% 
  summarise(n = n()) %>% 
  group_by(x) %>% 
  mutate(p_y_given_do_x = n / sum(n))

```

```{r data_a_table, echo=FALSE}
part_a %>% 
  kable() %>% 
  kable_styling(bootstrap_options = c('hover', 'striped', 'responsive'))
```


## Part b

To simulate observational data, we need to include the dependence of $X$ on $Z$.

```{r data_b}
N <- 100000  # number of individuals

set.seed(95400)

p_x_y_z <- tibble(
    id = 1:N,
    
    z = rbinom(N, 1, r),
    x = rbinom(N, 1, if_else(z == 0, q0, q1)),
    
    p_y_given_x_z = case_when(
      x == 0 & z == 0 ~ p00,
      x == 0 & z == 1 ~ p01,
      x == 1 & z == 0 ~ p10,
      x == 1 & z == 1 ~ p11
    ),
    
    y = rbinom(N, 1, p_y_given_x_z)
  ) %>% 
  group_by(x, y, z) %>% 
  count() %>% 
  ungroup() %>% 
  mutate(p = n / sum(n))

```


```{r p_x_y_z_table, echo=FALSE}
p_x_y_z %>% 
  kable() %>% 
  kable_styling(bootstrap_options = c('hover', 'striped', 'responsive'))
```

In order to apply the causal effect rule, we'll need $\mathbb P(x \mid z)$.

```{r p_x_given_z}
p_x_given_z <- p_x_y_z %>% 
  group_by(x, z) %>% 
  summarise(n = sum(n)) %>% 
  group_by(z) %>% 
  mutate(p = n / sum(n)) %>% 
  ungroup()
```


```{r p_x_given_z_table, echo=FALSE}
p_x_given_z %>% 
  kable() %>% 
  kable_styling(bootstrap_options = c('hover', 'striped', 'responsive'))
```

We can then add the conditional probabilities to the joint distribution table, then sum overal all the $Z$ variables.

```{r p_y_given_do_x}
p_y_given_do_x <- p_x_y_z %>% 
  inner_join(
    p_x_given_z, 
    by = c('x', 'z'), 
    suffix = c('_num', '_denom')
  ) %>% 
  mutate(p = p_num / p_denom) %>% 
  group_by(x, y) %>%
  summarise(p = sum(p)) 
```

```{r p_y_given_do_x_table, echo=FALSE}
p_y_given_do_x %>% 
  kable() %>% 
  kable_styling(bootstrap_options = c('hover', 'striped', 'responsive'))
```

The causal effect estimates are very close to the simulated intervention.

## Part c

We can calculate ACE simply by taking the difference of the causal effect estimates.

```{r ace}
ace <- p_y_given_do_x %>% 
  spread(x, p) %>% 
  filter(y == 1) %>% 
  mutate(ace = `1` - `0`) %>% 
  pull(ace)

ace
```

This is different from the overall probability differences.

```{r risk_difference}
p_y_given_x <- p_x_y_z %>% 
  group_by(x, y) %>% 
  summarise(n = sum(n)) %>% 
  group_by(x) %>% 
  mutate(p = n / sum(n)) %>% 
  select(-n) 

risk_difference <- p_y_given_x %>% 
  spread(x, p) %>% 
  filter(y == 1) %>% 
  mutate(rd = `1` - `0`) %>% 
  pull(rd)

risk_difference
```

Making $X$ independent of $Z$ would minimise the disrepancy between ACE and RD, which would turn the adjustment formula into the formulat for $\mathbb P(y \mid x$. In other words, setting $q_0 = q_1 = \mathbb P(X = 1)$ would do the trick.

## Part d

Note that the desegregated causal effects

* $p_{1, 0} - p_{0, 0}$ is `r p10 - p00`; and 
* $p_{1, 1} - p_{0, 1}$ is `r p11 - p01`,

are both consisent with our calculation for the overall causal effect, ACE = `r percent(ace)`. The generated data are an illustration of Simpson's reversal because the risk difference, `r percent(risk_difference)`, has the opposite sign.
