---
title: "BDA3 Chapter 3 Exercise 11"
author: "Brian Callander"
date: "2018-10-21"
tags: bda chapter 3, solutions, bayes
always_allow_html: yes
output: 
  md_document:
    variant: markdown
    preserve_yaml: yes
---

Here's my solution to exercise 11, chapter 3, of [Gelman's](https://andrewgelman.com/) *Bayesian Data Analysis* (BDA), 3rd edition. There are [solutions](http://www.stat.columbia.edu/~gelman/book/solutions.pdf) to some of the exercises on the [book's webpage](http://www.stat.columbia.edu/~gelman/book/).

<!--more-->

<div style="display:none">
  $\DeclareMathOperator{\dbinomial}{Binomial}
   \DeclareMathOperator{\dbern}{Bernoulli}
   \DeclareMathOperator{\dpois}{Poisson}
   \DeclareMathOperator{\dnorm}{Normal}
   \DeclareMathOperator{\dt}{t}
   \DeclareMathOperator{\dcauchy}{Cauchy}
   \DeclareMathOperator{\dexponential}{Exp}
   \DeclareMathOperator{\duniform}{Uniform}
   \DeclareMathOperator{\dgamma}{Gamma}
   \DeclareMathOperator{\dinvgamma}{InvGamma}
   \DeclareMathOperator{\invlogit}{InvLogit}
   \DeclareMathOperator{\logit}{Logit}
   \DeclareMathOperator{\ddirichlet}{Dirichlet}
   \DeclareMathOperator{\dbeta}{Beta}$
</div>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  error = TRUE,
  # cache = TRUE,
  dev = "svglite",
  fig.ext = ".svg" 
)

library(tidyverse)
library(scales)
library(kableExtra)

library(rstan)
library(rstanarm)
library(tidybayes)

rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

theme_set(theme_bw())

```

```{r data_generation, include = FALSE}
tibble(
    dose_log_g_ml = c(-0.86, -0.3, -0.05, 0.73),
    animals = c(5, 5, 5, 5),
    deaths = c(0, 1, 3, 5)
  ) %>% 
  write_csv('data/chapter_03_exercise_11.csv')

```

We will analyse [the data](data/chapter_03_exercise_11.csv) given in section 3.7 using different priors.

```{r data}
df <- read_csv('data/chapter_03_exercise_11.csv') 
```

```{r data_table, echo = FALSE}
df %>% 
  kable() %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"))
```

Here is the model specification.

$$
\begin{align}
  y_i \mid \theta_i 
  &\sim 
  \dbinomial(n_i, \theta_i)
  \\
  \logit(\theta_i)
  &=
  \alpha + \beta x_i
  \\
  \alpha
  &\sim
  \dnorm(0, 2^2)
  \\
  \beta
  &\sim
  \dnorm(10, 10^2)
\end{align}
$$

We won't use a grid approximation to the posterior but instead just use [Stan](https://www.rdocumentation.org/packages/rstanarm/versions/2.17.4/topics/stan_glm) because it is a lot simpler.

```{r model}
m <- rstanarm::stan_glm(
  cbind(deaths, animals - deaths)  ~ 1 + dose_log_g_ml,
  family = binomial(link = logit),
  data = df,
  prior_intercept = normal(0, 2),
  prior = normal(10, 10),
  warmup = 500,
  iter = 4000
)

summary(m)
```

The [tidybayes package](https://mjskay.github.io/tidybayes/articles/tidybayes.html) offers convenient functions for drawing from the posterior. We'll also add in our `LD50` estimate.

```{r draws}
draws <- m %>% 
  tidybayes::spread_draws(`(Intercept)`, dose_log_g_ml) %>% 
  rename(
    alpha = `(Intercept)`,
    beta = dose_log_g_ml
  ) %>% 
  mutate(LD50 = -alpha / beta)
```

```{r joint_posterior_plot, echo = FALSE}
n_draws <- max(draws$.draw)

draws %>% 
  ggplot() +
  aes(alpha, beta) +
  stat_density_2d(aes(fill = ..level..), geom = 'polygon') +
  scale_y_continuous(limits = c(0, NA)) +
  labs(
    title = 'Join posterior distribution of α and β',
    subtitle = str_glue('{comma(n_draws)} draws'),
    x = 'α',
    y = 'β'
  )
```

The estimates look much the same with the more informative priors as with the uninformative priors. The posterior probability that $\beta > 0$ is:

```{r}
draws %>% 
  mutate(positive = beta > 0) %>% 
  summarise(mean(positive)) %>% 
  pull() %>% 
  percent()

```

The posterior LD50 estimate (conditional on $\beta > 0$) is as follows:

```{r}
draws %>% 
  filter(beta > 0) %>% 
  ggplot() + 
  aes(LD50) +
  geom_histogram(bins = 50) +
  geom_vline(xintercept = 0, linetype = 'dashed', colour = 'chocolate') +
  labs(
    y = 'Count',
    title = 'Histogram of posterior LD50 estimate',
    subtitle = 'Conditional on β > 0'
  )
```

