---
title: "BDA3 Chapter 3 Exercise 2"
author: "Brian Callander"
date: "2018-09-15"
tags: bda chapter 3, solutions, bayes, multinomial, dirichlet, beta
tldr: Here's my solution to exercise 2, chapter 3, of Gelman's Bayesian Data Analysis (BDA), 3rd edition.
always_allow_html: yes
output: 
  md_document:
    variant: markdown
    preserve_yaml: yes
---

Here's my solution to exercise 2, chapter 3, of [Gelman's](https://andrewgelman.com/) *Bayesian Data Analysis* (BDA), 3rd edition. There are [solutions](http://www.stat.columbia.edu/~gelman/book/solutions.pdf) to some of the exercises on the [book's webpage](http://www.stat.columbia.edu/~gelman/book/).

<!--more-->

<div style="display:none">
  $\DeclareMathOperator{\dbinomial}{Binomial}
   \DeclareMathOperator{\dbern}{Bernoulli}
   \DeclareMathOperator{\dpois}{Poisson}
   \DeclareMathOperator{\dnorm}{Normal}
   \DeclareMathOperator{\dcauchy}{Cauchy}
   \DeclareMathOperator{\dexponential}{Exp}
   \DeclareMathOperator{\dgamma}{Gamma}
   \DeclareMathOperator{\dinvgamma}{InvGamma}
   \DeclareMathOperator{\invlogit}{InvLogit}
   \DeclareMathOperator{\logit}{Logit}
   \DeclareMathOperator{\ddirichlet}{Dirichlet}
   \DeclareMathOperator{\dbeta}{Beta}$
</div>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  error = TRUE,
  dev = "svglite",
  fig.ext = ".svg" 
)

library(tidyverse)
library(scales)
library(kableExtra)

theme_set(theme_bw())

```

```{r data, echo = FALSE}
df <- tibble(
  survey = c('pre-debate', 'post-debate'),
  bush = c(294, 288),
  dukakis = c(307, 332),
  other = c(38, 19),
  total = 639
)

df %>% 
  kable() %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"))
```

Let $\theta_\text{pre} \sim \ddirichlet_3(\alpha_\text{pre})$ and $\theta_\text{post} \sim \ddirichlet_3(\alpha_\text{post})$ be the pre- and post-debate priors. Then the posteriors are 

$$
\begin{align}
  \theta_\text{pre} \mid y &\sim \ddirichlet\left(294 + \alpha_{(\text{pre}, \text{b})}, 307 + \alpha_{(\text{pre}, \text{d})}, 38 + \alpha_{(\text{pre}, \text{o})} \right)
  \\
  \theta_\text{post} \mid y &\sim \ddirichlet\left(288 + \alpha_{(\text{post}, \text{b})}, 332 + \alpha_{(\text{post}, \text{d})}, 19 + \alpha_{(\text{post}, \text{o})} \right)
.
\end{align}
$$

Denote the proportion of support for Bush amongst Bush/Dukakis supporters by $\phi_\bullet := \frac{\theta_{(\bullet, \text{b})}}{\theta_{(\bullet, \text{b})} + \theta_{(\bullet, \text{d})}}$. The results of the [previous exercise](./chapter_03_exercise_01.html) show that we can treat this as a beta-binomial problem by simply ignoring the results for `other`. More precisely,

$$
\begin{align}
  \phi_\text{pre} \mid y &\sim \dbeta\left(294 + \alpha_{(\text{pre}, \text{b})}, 307 + \alpha_{(\text{pre}, \text{d})} \right)
  \\
  \phi_\text{post} \mid y &\sim \dbeta\left(288 + \alpha_{(\text{post}, \text{b})}, 332 + \alpha_{(\text{post}, \text{d})} \right)
\end{align}
.
$$

Let's plot these posteriors and their difference $\delta := (\phi_\text{post} \mid y) - (\phi_\text{pre} \mid y)$.

```{r posteriors}
alpha_bush <- 1
alpha_dukakis <- 1

posteriors <- expand.grid(
    value = seq(0, 1, 0.001),
    survey = df$survey
  ) %>% 
  as_tibble() %>% 
  inner_join(df, by = 'survey') %>% 
  mutate(
    post_bush = bush + alpha_bush,
    post_dukakis = dukakis + alpha_dukakis,
    density = dbeta(value, post_bush, post_dukakis)
  ) 
```

```{r posteriors_plot, echo = FALSE}
posteriors %>% 
  ggplot() +
  aes(value, density, colour = survey) +
  geom_line() +
  scale_x_continuous(limits = c(0.3, 0.7), labels = percent) +
  labs(
    x = 'Proportion in support of Bush',
    y = 'Density',
    title = 'Posterior estimate of proportion in support of Bush',
    subtitle = 'Pre- vs. post-debate',
    colour = 'Survey'
  )
  
```


```{r shift}
N <- 10000

shift <- expand.grid(
    draw = 1:N, 
    survey = df$survey
  ) %>% 
  as_tibble() %>% 
  inner_join(df, by = 'survey') %>% 
  mutate(
    post_bush = bush + alpha_bush,
    post_dukakis = dukakis + alpha_dukakis,
    value = rbeta(n(), post_bush, post_dukakis)
  ) %>% 
  select(draw, survey, value) %>% 
  spread(survey, value) %>% 
  mutate(difference = `post-debate` - `pre-debate`) 
```

```{r shift_plot, echo = FALSE}
shift %>% 
  ggplot() +
  aes(difference) +
  geom_histogram(bins = 50) +
  geom_vline(xintercept = 0, linetype = 'dashed', colour = 'dark orange') +
  scale_x_continuous(labels = percent) +
  labs(
    x = 'Percentage point shift',
    y = 'Count',
    title = 'Shift in support for Bush vs Dukakis',
    subtitle = str_glue("{comma(N)} draws from the posterior of the pre- vs. post-debate difference")
  )
  
```

There is a `r percent(mean(shift$difference > 0))` probability (the area above 0) that there was a positive shift towards Bush.