---
title: "BDA3 Chapter 2 Exercise 12"
author: "Brian Callander"
date: "2018-09-01"
tags: bda chapter 2, solutions, bayes, poisson, jeffrey prior, gamma
tldr: Here's my solution to exercise 12, chapter 2, of Gelman's Bayesian Data Analysis (BDA), 3rd edition.
always_allow_html: yes
output: 
  md_document:
    variant: markdown
    preserve_yaml: yes
---

Here's my solution to exercise 12, chapter 2, of [Gelman's](https://andrewgelman.com/) *Bayesian Data Analysis* (BDA), 3rd edition. There are [solutions](http://www.stat.columbia.edu/~gelman/book/solutions.pdf) to some of the exercises on the [book's webpage](http://www.stat.columbia.edu/~gelman/book/).

<!--more-->

<div style="display:none">
  $\DeclareMathOperator{\dbinomial}{binomial}
   \DeclareMathOperator{\dbern}{Bernoulli}
   \DeclareMathOperator{\dnorm}{normal}
   \DeclareMathOperator{\dcauchy}{Cauchy}
   \DeclareMathOperator{\dgamma}{gamma}
   \DeclareMathOperator{\invlogit}{invlogit}
   \DeclareMathOperator{\logit}{logit}
   \DeclareMathOperator{\dbeta}{beta}$
</div>


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  error = TRUE,
  cache = TRUE
)

library(tidyverse)
library(kableExtra)

theme_set(theme_bw())

```

Suppose $\theta$ has a Poisson likelihood so that $\log p(y \mid \theta) \propto y \log(\theta) - \theta$. We will find Jeffrey's prior for $\theta$ and the gamma distribution that most closely approximates it.

The derivative of the log likelihood is $\frac{y}{\theta} - 1$ and the second derivative is $-\frac{y}{\theta^2}$. It follows that the Fisher information for $\theta$ is

$$
J(\theta)
=
\mathbb E \left( \frac{y}{\theta^2} \right)
=
\frac{1}{\theta}
,
$$

so Jeffrey's prior is $p(\theta) \propto \frac{1}{\sqrt{\theta}}$. This is an improper prior because

$$
\int_0^\infty \theta^{-\frac{1}{2}} d\theta
=
\left[ 2\theta^{\frac{1}{2}} \right]_0^\infty
=
\infty
.
$$

Since Jeffrey's prior is improper, we can try approximate it with a gamma prior. Let $\alpha, \beta \in (0, \infty)$ be the shape and rate parameters of a gamma distribution.  Then

$$
\dgamma(\theta \mid \alpha, \beta) 
\propto
x^{\alpha - 1}e^{-\beta x}
.
$$

Choosing $\alpha = \frac{1}{2}$ and $\beta = 0$ yields Jeffrey's prior. However, $\beta$ must be positive for the gamma distribution to be proper, so we can choose $\beta = \epsilon$ sufficiently small. We'll use the [smallest positive float](https://stat.ethz.ch/R-manual/R-devel/library/base/html/zMachine.html) representable in R.

```{r gamma_jeffrey}
epsilon <- .Machine$double.xmin

upper_limit <- 100000000
step <- upper_limit / 10000

prior <- tibble(theta = seq(0, upper_limit, step)) %>% 
  mutate(density = dgamma(theta, 0.5, epsilon)) 
```

```{r gamma_jeffrey_plot, echo = FALSE}
prior %>% 
  ggplot() +
  aes(theta, density) +
  geom_area(fill = 'skyblue') +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = function(x) signif(x, 2), limits = c(0, NA)) +
  labs(
    x = 'θ',
    y = 'Density',
    title = str_glue("Gamma(0.5, {signif(epsilon, digits = 2)}) as an approximation of Jeffrey's Poisson prior"),
    subtitle = str_glue(paste(
      'Truncated above at {scales::comma(upper_limit)}',
      'Calculated in steps of {scales::comma(step)}',
      sep = '\n'
    ))
  )

```

