---
title: "BDA3 Chapter 3 Exercise 4"
author: "Brian Callander"
date: "2018-10-05"
tags: bda chapter 3, solutions, bayes, odds ratio, noninformative priors
always_allow_html: yes
output: 
  md_document:
    variant: markdown
    preserve_yaml: yes
---

Here's my solution to exercise 4, chapter 3, of [Gelman's](https://andrewgelman.com/) *Bayesian Data Analysis* (BDA), 3rd edition. There are [solutions](http://www.stat.columbia.edu/~gelman/book/solutions.pdf) to some of the exercises on the [book's webpage](http://www.stat.columbia.edu/~gelman/book/).

<!--more-->

<div style="display:none">
  $\DeclareMathOperator{\dbinomial}{Binomial}
   \DeclareMathOperator{\dbern}{Bernoulli}
   \DeclareMathOperator{\dpois}{Poisson}
   \DeclareMathOperator{\dnorm}{Normal}
   \DeclareMathOperator{\dt}{t}
   \DeclareMathOperator{\dcauchy}{Cauchy}
   \DeclareMathOperator{\dexponential}{Exp}
   \DeclareMathOperator{\dgamma}{Gamma}
   \DeclareMathOperator{\dinvgamma}{InvGamma}
   \DeclareMathOperator{\invlogit}{InvLogit}
   \DeclareMathOperator{\logit}{Logit}
   \DeclareMathOperator{\ddirichlet}{Dirichlet}
   \DeclareMathOperator{\dbeta}{Beta}$
</div>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  error = TRUE,
  dev = "svglite",
  fig.ext = ".svg" 
)

library(tidyverse)
library(scales)
library(kableExtra)

theme_set(theme_bw())

```

Suppose we have two independent trials where the likelihood of death is binomial, $y_i \mid p_0, p_1 \sim \dbinomial(n_i, p_i)$, $i = 0, 1$. We will compare two different non-informative priors on the odds ratio

$$
\theta
:=
\frac{p_0}{1 - p_0} / \frac{p_1}{1 - p_1} 
.
$$

Here are the given data.

```{r data}
df <- tibble(
    cohort = c('control', 'treatment'),
    patients = c(674, 680),
    deaths = c(39, 22)
  ) %>% 
  mutate(survived = patients - deaths)
```

```{r data_table, echo = FALSE}
df %>% 
  kable() %>% 
  kable_styling()
```

We'll need a couple of functions for drawing random samples for $\theta$.

```{r functions}
odds <- function(p) p / (1 - p)

simulate <- function(n, k, a, b, draws = 10000)
  # draws from posterior
  # n bernouille trials, k successes with beta(a, b) prior
  tibble(
    draw = 1:draws, 
    value = rbeta(draws, k + a, n - k + b)
  )

posterior <- function(a, b, draws = 10000)
  # random samples from theta posterior with beta(a, b) prior
  df %>% 
    transmute(
      cohort,
      draws = map2(patients, deaths, simulate, a, b, draws)
    ) %>% 
    unnest(draws) %>% 
    spread(cohort, value) %>% 
    mutate(theta = odds(control) / odds(treatment))
```
  
Let's compare a uniform prior to a prior close to $\dbeta(0, 0)$.

```{r posteriors}
uni <- posterior(1, 1) %>% 
  mutate(prior = 'uniform')

zero <- posterior(0.000000001, 0.000000001) %>% 
  mutate(prior = 'zero')

posteriors <- bind_rows(uni, zero)
```

```{r posteriors_plot, echo = FALSE}
posteriors %>% 
  ggplot() +
  aes(theta, fill = prior) +
  geom_histogram(position = 'identity', alpha = 0.5, bins = 50) +
  labs(
    x = 'θ',
    y = 'Count',
    title = 'Comparison of non-informative priors on the posterior odds ratio'
  )
```

Here are the 95% posterior credible intervals for $\theta$.

```{r posteriors_quantiles}
cis <- posteriors %>% 
  group_by(prior) %>% 
  summarise(
    q05 = quantile(theta, 0.05),
    q50 = quantile(theta, 0.5),
    q95 = quantile(theta, 0.95)
  )
```

```{r posteriors_quantiles_table, echo = FALSE}
cis %>% 
  kable() %>% 
  kable_styling()
```

The estimates with the "zero" prior are slightly higher than those from the uniform prior, especially in the tails.

