---
title: "BDA3 Chapter 5 Exercise 15"
author: "Brian Callander"
date: "2019-02-05"
tags: bda chapter 5, solutions, normal, hierarchical model, posterior predictive distribution
tldr: "Here's my solution to exercise 15, chapter 5, of Gelman's Bayesian Data Analysis (BDA), 3rd edition."
always_allow_html: yes
output: 
  md_document:
    variant: markdown
    preserve_yaml: yes
---

Here's my solution to exercise 15, chapter 5, of [Gelman's](https://andrewgelman.com/) *Bayesian Data Analysis* (BDA), 3rd edition. There are [solutions](http://www.stat.columbia.edu/~gelman/book/solutions.pdf) to some of the exercises on the [book's webpage](http://www.stat.columbia.edu/~gelman/book/).

<!--more-->

<div style="display:none">
  $\DeclareMathOperator{\dbinomial}{Binomial}
   \DeclareMathOperator{\dbern}{Bernoulli}
   \DeclareMathOperator{\dpois}{Poisson}
   \DeclareMathOperator{\dnorm}{Normal}
   \DeclareMathOperator{\dt}{t}
   \DeclareMathOperator{\dcauchy}{Cauchy}
   \DeclareMathOperator{\dexponential}{Exp}
   \DeclareMathOperator{\duniform}{Uniform}
   \DeclareMathOperator{\dgamma}{Gamma}
   \DeclareMathOperator{\dinvgamma}{InvGamma}
   \DeclareMathOperator{\invlogit}{InvLogit}
   \DeclareMathOperator{\logit}{Logit}
   \DeclareMathOperator{\ddirichlet}{Dirichlet}
   \DeclareMathOperator{\dbeta}{Beta}$
</div>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  error = TRUE,
  # cache = TRUE,
  dev = "svglite"
)

library(tidyverse)
library(scales)
library(kableExtra)

library(rstan)
library(tidybayes)

rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

theme_set(theme_bw())

```

The [data provided](http://www.stat.columbia.edu/~gelman/book/data/meta.asc) are in an awkward format. I've [downloaded it](data/chapter_05_exercise_15_table_5.4.txt) with minor modifications to make it easier to parse.

```{r data}
df <- read_delim(
    'data/chapter_05_exercise_15_table_5.4.txt', 
    delim=' ',
    skip=4,
    trim_ws=TRUE
  ) %>% 
  transmute(
    study,
    y = log(treated.deaths / (treated.total - treated.deaths)) - log(control.deaths / (control.total - control.deaths)),
    sigma2 = (1 / treated.deaths) + (1 / (treated.total - treated.deaths)) + (1 / control.deaths) + (1 / (control.total - control.deaths)),
    sigma = sqrt(sigma2),
    n = treated.total + control.total
  ) %>% 
  select(-sigma2)

```

```{r data_table, echo=FALSE}
df %>% 
  kable(caption = "The meta-analysis data from table 5.4, page 124.") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"))

```

We'll use the model described in the book. Note that by not explicitly giving a prior for $\mu$ or $\tau$, stan gives them a uniform prior.

```{r model, results='hide'}
model <- rstan::stan_model('src/ex_05_15.stan')
```

```{r model_code, echo=FALSE}
model
```

Let's fit the model.

```{r fit}
set.seed(57197)

fit <- model %>% 
  sampling(
    data = tidybayes::compose_data(df),
    warmup = 1000,
    iter = 5000
  )
```

We'll draw the posterior population parameters separately from the study parameters purely for convenience.

```{r draws}
pop_params <- fit %>% 
  tidybayes::spread_draws(mu, tau)

draws <- fit %>% 
  tidybayes::spread_draws(mu, tau, theta[study])
```


The population standard deviation $\tau$ has most of its mass below 0.4.

```{r tau_plot, echo=FALSE, fig.cap="A histogram of the posterior draws for τ."}
pop_params %>% 
  ggplot() +
  aes(tau) +
  geom_histogram() +
  labs(
    x = 'τ',
    y = 'Count',
    title = 'Posterior of τ'
  )
```

As in figure 5.6, the effect estimates are almost identical for $\tau \approx 0$ and spread out as $\tau$ increases.

```{r theta_vs_tau, echo=FALSE, fig.cap="Conditional posterior means of treatment effects E(θ | τ, y), as functions of the between-study standard deviation τ."}
draws %>% 
  ggplot() +
  aes(tau, theta, colour = ordered(study)) +
  geom_smooth(se=FALSE) +
  labs(
    x = 'τ',
    y = 'θ',
    title = 'Estimated effect as a function of the standard deviation',
    subtitle = 'We use smooths of the posterior draws',
    colour = 'Study'
  )
  
```

Let's get the median effect estimates with 95% posterior intervals. The low sample size estimates (dark dots) remain close to the population mean ($\mu$), whereas the larger sample size estimates (light dots) can move closer to the unpooled estimates (dotted line).

```{r cis}
cis <- draws %>% 
  median_qi() %>% 
  inner_join(df, by = 'study')
```


```{r cis_plot, echo=FALSE, fig.cap="Comparison of the crude effect estimates with the posterior median effects."}
cis %>% 
  ggplot() +
  aes(x = y) +
  geom_abline(slope = 1, intercept = 0, linetype = 'dotted', colour = 'black') +
  geom_ribbon(aes(ymin = mu.lower, ymax = mu.upper), alpha = 0.3, colour = NA, fill = 'chocolate') +
  geom_hline(yintercept = cis$mu, linetype = 'dashed', colour = 'chocolate') +
  geom_pointrange(aes(x = y, ymin = theta.lower, y = theta, ymax = theta.upper, colour = log(n))) +
  labs(
    x = 'Crude effect estimate, y',
    y = 'Posterior median effect, θ',
    title = 'Crude effect estimates vs. posterior median effect estimates',
    subtitle = paste(
      'The orange band is a 95% credible interval for μ',
      'The dotted line corresponds to the unpooled estimates',
      sep = '\n'
    ),
    colour = 'log(sample size)'
  )
```

To estimate an effect for a new study, we draw $\theta \sim \dnorm(\mu, \tau)$.

```{r new_theta}
new_theta <- pop_params %>% 
  mutate(theta = rnorm(n(), mu, tau)) 
```

```{r new_study_positive}
prob_new_theta_positive <- new_theta %>% 
  summarise(sum(theta > 0) / n()) %>% 
  pull() 
  
```

It has a `r prob_new_theta_positive %>% percent()` probability of being positive.

```{r new_theta_plot, echo=FALSE, fig.cap='Posterior simulations of a new treatment effect.'}
new_theta %>% 
  ggplot() +
  aes(theta) +
  geom_histogram() +
  geom_vline(xintercept = 0, linetype = 'dashed', colour = 'chocolate') +
  labs(
    x = 'θ',
    y = 'Count',
    title = 'Posterior simulations of a new treatment effect',
    subtitle = str_glue('There is a {percent(prob_new_theta_positive)} posterior probability that θ > 0')
  )
  

```

